Fast and optimal centroiding of faint stars

### 2014-12-11 ms/figure.tex
update method

\title{Towards fast and optimal centroiding of faint stars}

@@

\section{Model}\label{sec:model}

In order to find the centroid of a star, we fit a simple 2d polynomial $P(x,y)=a+bx+cy+dx^2+exy+fy^2$ to the $3\times3$ patch centered on the brightest pixel of the
image. We apply this method to star images in the sky-dominated regime. In this regime, the covariance matrix of pixel uncertianties is simply given by $\sigma^{2}\mathbf{I}$, where $\sigma^{2}$ is variance of the uncorrelated Gaussian noise, and $\mathbf{I}$ is a 9$\times$9 identity matrix. Upon constructing a universal 9$\times$6 design matrix

\begin{equation}
    \mathbf{A} = 
    \begin{bmatrix}
        1 & x_{1} & y_{1} & x_{1}^{2} & x_{1}y_{1} & y_{1}^{2} \\
        . & . & . & . & . & .  \\
        . & . & . & . & . & .  \\
        . & . & . & . & . & .  \\
        1 & x_{9} & y_{9} & x_{9}^{2} & x_{9}y_{9} & y_{9}^{2}
    \end{bmatrix},
\end{equation}
the free parameters of the model $\{a,b,c,d,e,f\}$ (hereafter compactly denoted by $\mathbf{X}$) can be determined by 
\beq
\mathbf{X} = (\mathbf{A}^{T}\mathbf{A})^{-1}\mathbf{A}^{T}\mathbf{Z},
\label{linearfit}
\eeq
where $\mathbf{Z}$ is given by $(I_{1},...,I_{9})^{T}$, with $I_{i}$, being the brightness of the $i-$th pixel of the 3$\times$3 patch. Afterwards, the best fit parameters can be used to compute the centroid coordinate

@@

Now, we modify the simple polynomial method explained above, by smoothing the image of the star with a Gaussian kernel with a given width (2 pixels in this study). Then we apply the same polynomial method to the 3$\times$3 patch around the brightest pixel of the smoothed image. The important difference between the previous approach and the modified method is that, there are non-zero covariances among the uncertainties of the brightness of different pixels of the smoothed image. That is, the covariance matrix $\mathbf{C}$ is non-diagonal. It can be shown that after smoothing, elements of the covariance matrix are given by
\begin{eqnarray}
\mathbf{C}_{ij} &=& \frac{\sigma^{2}}{4\pi w^{2}} \exp(-\frac{r_{ij}^{2}}{4\pi w^{2}}),\\
r_{ij}^{2} &=& (x_{i} - x_{j})^{2} + (y_{i} - y_{j})^{2},
\label{nondiagonal}
\end{eqnarray}
where $w$ is width of the kernel, and $x_{i}$ , $y_{i}$ are the $x$, and $y$ coordinates of the $i$-th pixel respectively. In the presence of the non-diagonal covariance matrix $\mathbf{C}$, equation (\ref{linearfit}) needs to be modified in the following way
\beq
\mathbf{X} = [\mathbf{A}^{T}\mathbf{C}^{-1}\mathbf{A}]^{-1}[\mathbf{A}^{T}\mathbf{C}^{-1}\mathbf{Z}].
\label{linearfit2}
\eeq

Therefore, for a given star and a smoothing kernel, the outcome of equation (\ref{linearfit2}) can be plugged into equation (\ref{center}) to find the centroid estimate of the star. Hereafter, we call this method the modified polynomial method.

Now, let us give a brief overview of centroiding method implemented in SDSS photometric pipeline \citep{sdss}: Consider any row or column of the 3$\times$3 patch around the brightest pixel of the smooth image and let us denote the brightness of these three pixels by $\{I_{-},I_{0},I_{+}\}$. The fundamental assumption behind the technique is that these brightness values can be approximated by the function $P(x) = Q\exp((x-x_{c})^{2}/2\beta^{2})$ evaluated at $x=-1,0,+1$; with $x_{c}$ being the position of centroid. Knowing the value of $P$ at $\{-1,0,1\}$, expanding $P(x)$ up to fourth order in $x$ leads to
\beq
x = \frac{s}{t}(1+k\frac{t}{4Q}(1-(s/t)^{2})),
\eeq
where k = 1.33, and

>>>>

\title{Fast and optimal centroiding of faint stars}

@@

\section{Methods}\label{sec:method}

\subsection{3$\times$3 polynomial method}

In order to find the centroid of a star, 
we fit a simple 2d polynomial $P(x,y)=a+bx+cy+dx^2+exy+fy^2$ to the
$3\times3$ patch centered on the brightest pixel of the
image. We apply this method to star images in the sky-dominated regime.
In this regime, the covariance matrix of pixel uncertianties is simply
given by $\sigma^{2}\mathbf{I}$, where $\sigma^{2}$ is variance of the
uncorrelated Gaussian noise, and $\mathbf{I}$ is a 9$\times$9 identity matrix.
Upon constructing a universal 9$\times$6 design matrix
\begin{equation}
    \mathbf{A} = 
    \begin{bmatrix}
        1 & x_{1} & y_{1} & x_{1}^{2} & x_{1}y_{1} & y_{1}^{2} \\
        . & . & . & . & . & .  \\
        . & . & . & . & . & .  \\
        . & . & . & . & . & .  \\
        1 & x_{9} & y_{9} & x_{9}^{2} & x_{9}y_{9} & y_{9}^{2}
    \end{bmatrix},
\end{equation}
the free parameters of the model $\{a,b,c,d,e,f\}$
(hereafter compactly denoted by $\mathbf{X}$) can be determined by 
\beq
\mathbf{X} = (\mathbf{A}^{T}\mathbf{A})^{-1}\mathbf{A}^{T}\mathbf{Z},
\label{linearfit}
\eeq
where $\mathbf{Z}$ is given by $(z_{1},...,z_{9})^{T}$,
with $z_{i}$, being the brightness of the $i-$th pixel of the 3$\times$3 patch.
Afterwards, the best fit parameters can be used to compute the centroid coordinate

@@

It is important to note that the algebraic operation in (\ref{center}) involves 
inverting a 2$\times$2 curvature matrix
\beq
  D = 
  \begin{bmatrix}
      2d & e\\
      e & 2f\\
  \end{bmatrix}.
\eeq
When D has a zero (or very close to zero) deteminant,
centroid estimated by (\ref{center}) gets arbitrarily 
large in which case the centroiding method leads to a catastrophic 
outlier. In order to tackle this issue, we add a small regularization term
proportional to $\sigma$ to the diagonal elements of the curvature matrix.
This soft regularization term prevents the determinant of the curvature matrix from
getting zero, and removes the catastrophic outliers.

\subsection{3$\times$3 modified polynomial method}
Now, we modify the simple polynomial method explained above,
by smoothing the image of the star with a Gaussian kernel
\beq
k(x) = \frac{1}{2\pi w^2}\exp(-\mathbf{x}^{2}/2w^{2}).
\eeq
Throughout this study, we keep the width of the Gaussian kernel $w$ constant.
Then we apply the same polynomial method to the 3$\times$3 patch around the brightest
pixel of the smoothed image. The important difference between the previous
approach and the modified method is that, there are non-zero covariances
among the uncertainties of the brightness of different pixels of the
smoothed image. That is, the covariance matrix $\mathbf{C}$ is non-diagonal.
It can be shown that after smoothing, elements of the covariance matrix are given by
\begin{eqnarray}
\mathbf{C}_{ij} &=& \frac{\sigma^{2}}{4\pi w^{2}} \exp(-\frac{r_{ij}^{2}}{4\pi w^{2}}),\\
r_{ij}^{2} &=& (x_{i} - x_{j})^{2} + (y_{i} - y_{j})^{2},
\label{nondiagonal}
\end{eqnarray}
where $w$ is width of the kernel, $\sigma$ is the per-pixel uncertainty, and $x_{i}$
, $y_{i}$ are the $x$, and $y$ coordinates of the $i$-th pixel
respectively. In the presence of the non-diagonal covariance matrix
$\mathbf{C}$, equation (\ref{linearfit}) needs to be modified in the following way
\beq
\mathbf{X} = [\mathbf{A}^{T}\mathbf{C}^{-1}\mathbf{A}]^{-1}[\mathbf{A}^{T}\mathbf{C}^{-1}\mathbf{Z}].
\label{linearfit2}
\eeq
Therefore, for a given star and a smoothing kernel,
the outcome of equation (\ref{linearfit2}) can be
plugged into equation (\ref{center}) to find the centroid estimate
of the star.
\subsection{3$\times$3 quartic approximation method}
Consider any row or column of the 3$\times$3 patch around
the brightest pixel of the smooth image and let us denote
the brightness of these three pixels by $\{I_{-},I_{0},I_{+}\}$.
The fundamental assumption behind the technique is that these
brightness values can be approximated
by the function $P(x) = Q\exp((x-x_{c})^{2}/2\beta^{2})$
evaluated at $x=-1,0,+1$; with $x_{c}$ being the position of centroid
of the three pixels. Knowing the value of $P$ at $\{-1,0,1\}$,
expanding $P(x)$ up to fourth order in $x$ leads to
\beq
x = \frac{s}{t}(1+k\frac{t}{4Q}(1-(s/t)^{2})),
\eeq
where k = 1.33, and
